A nice acyclic matching on the nerve of the partition lattice

### 2020-09-29 paper3/equivariant.tex
paper3v4

\documentclass{elsarticle}
\usepackage{amsmath,amsthm}
\usepackage{bbm}
\newtheorem{df}{Definition}[section]
\newtheorem{prop}[df]{Proposition}
\newtheorem{lem}[df]{Lemma}
\newtheorem{thm}[df]{Theorem}
\newtheorem{cor}[df]{Corollary}
\newtheorem{ex}[df]{Example}
\newtheorem{rem}[df]{Remark}
\def\N{\mathbbm N}
\def\Z{\mathbbm Z}
\def\C{\mathbbm C}
\def\id{\operatorname{id}}
%\def\card{\operatorname{card}}
\journal{arXiv.org}
\begin{document}
\begin{frontmatter}
\title{A nice acyclic matching on the nerve of the partition lattice}
\author[ruelle]{Ralf Donau}
\address[ruelle]{Fachbereich Mathematik, Universit\"at Bremen, Bibliothekstra\ss e 1, 28359 Bremen, Germany}
\ead{ruelle@math.uni-bremen.de}
\begin{abstract}
The author has already proven that the space $\Delta(\Pi_n)/G$ is homotopy equivalent to a wedge of spheres of dimension $n-3$ for all natural numbers $n\geq 3$ and all subgroups $G\subset S_1\times S_{n-1}$. We wish to construct an acyclic matching on $\Delta(\Pi_n)/G$ that allows us to give a basis of its cohomology. This is also a more elementary approach to determining the number of spheres. Furthermore we give a description of the group action by an action on the spheres. We also obtain another result that we call Equivariant Patchwork Theorem.
\end{abstract}
\begin{keyword}
Discrete Morse Theory\sep Regular trisp\sep Acyclic matching\sep Equivariant homotopy
\end{keyword}
\end{frontmatter}
\section{Introduction}
Let $n\geq 3$ and let $\Pi_n$ denote the poset consisting of all partitions of $[n]:=\{1,\dots,n\}$ ordered by refinement, such that the finer partition is the smaller partition. Let $\overline{\Pi}_n$ denote the poset obtained from $\Pi_n$ by removing both the smallest and greatest element, which are $\{\{1\},\dots,\{n\}\}$ and $\{[n]\}$, respectively. We consider $\overline{\Pi}_n$ as a category, which is acyclic, and define $\Delta(\overline{\Pi}_n)$ to the nerve of the acyclic category $\overline{\Pi}_n$, which is a regular trisp, see \cite[Chapter 10]{buch}. The symmetric group $S_n$ operates on $\Delta(\overline{\Pi}_n)$ in a natural way.

@@

In this paper we construct an $(S_1\times S_{n-1})$-equivariant acyclic matching on the face poset ${\cal F}(\Delta(\overline{\Pi}_n))$ of $\Delta(\overline{\Pi}_n)$ for $n\geq 3$ such that we have a description of the critical simplices. This induces an acyclic matching on $\Delta(\overline{\Pi}_n)/G$ for any subgroup $G\subset S_1\times S_{n-1}$. Another benefit of having a description of the critical simplices is that we can easily give cocycle representatives of the generators of the cohomology, which can be useful for further analysis of the cohomology.

@@

By Remark \ref{card} the cardinality of $C_n$ is $(n-1)!$ which equals the cardinality of $S_1\times S_{n-1}$. Hence the action is free.
\end{proof}
Let $G\subset S_1\times S_{n-1}$ be an arbitrary subgroup.
\begin{rem}
The cardinality of $C_n/G$ is the index of $G$ in $S_1\times S_{n-1}$.
\end{rem}
\begin{proof}
Apply Lemma \ref{freetrans}.
\end{proof}
Now Theorem \ref{rdthm} follows as a corollary. We can either apply Corollary \ref{moeglichkeit1} or Corollary \ref{moeglichkeit2}.
\begin{cor}
The topological space $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of spheres of dimension $n-3$. The number of spheres is the index of $G$ in $S_1\times S_{n-1}$.
\end{cor}
Now we consider the top cohomology $H^{n-3}(\Delta(\overline{\Pi}_n);\C)$ of $\Delta(\overline{\Pi}_n)$. For each simplex $\sigma\in C_n$ we obtain a $\sigma^*\in H^{n-3}(\Delta(\overline{\Pi}_n);\C)$ which is represented by a cocycle that maps $\sigma$ to $1$ and all other simplices to $0$. We obtain a basis $\{\sigma^*\mid\sigma\in C_n\}$ of the top cohomology which is called \emph{splitting basis} in \cite{wachs}.
\section*{Acknowledgments}
The author would like to thank Dmitry N. Kozlov for this interesting problem, Ragnar Freij and Giacomo d'Antonio for the helpful discussions.
\bibliographystyle{model1-num-names}
\begin{thebibliography}{00}
\bibitem{arone}
G. Arone, \textit{A branching rule for partition complexes}, \texttt{arXiv:1508.05086 [math.AT]} (2015).
\bibitem{donau2}
R. Donau, \textit{On a quotient topology of the partition lattice with forbidden block sizes}, Topology and its Applications \textbf{159} (8) (2012), pp. 2052-2057.
\bibitem{donau}
R. Donau, \textit{Quotients of the order complex $\Delta(\overline{\Pi}_n)$ by subgroups of the Young subgroup $S_1\times S_{n-1}$}, Topology and its Applications \textbf{157} (16) (2010), pp. 2476-2479.
\bibitem{torsion}
R. Donau, \textit{Quotients of the topology of the partition lattice which are not homotopy equivalent to wedges of spheres}, \texttt{arXiv:1202.4368 [math.AT]} (2012).
\bibitem{forman}
R. Forman, \textit{Morse theory for cell complexes}, Adv. Math. \textbf{134} (1) (1998), pp. 90-145.
\bibitem{freij}
R. Freij, \textit{Equivariant discrete Morse theory}, Discrete Mathematics \textbf{309} (12) (2009), pp. 3821-3829.
\bibitem{hatcher}
A. Hatcher, \textit{Algebraic Topology}, Cambridge University Press, 2008.
\bibitem{clmap}
D.N. Kozlov, \textit{Closure maps on regular trisps}, Topology and its Applications \textbf{156} (15) (2009), pp. 2491-2495.
\bibitem{buch}
D. Kozlov, \textit{Combinatorial Algebraic Topology}, Algorithms and Computation in Mathematics \textbf{21}, Springer-Verlag Berlin Heidelberg, 2008.
\bibitem{wachs}
M.L. Wachs, \textit{On the (co)homology of the partition lattice and the free Lie algebra}, Discrete Mathematics \textbf{193} (1-3) (1998), pp. 287-319.
\end{thebibliography}
\end{document}

>>>>

\documentclass{elsarticle}
\usepackage{amsmath,amsthm}
\usepackage{bbm}
\newtheorem{df}{Definition}[section]
\newtheorem{prop}[df]{Proposition}
\newtheorem{lem}[df]{Lemma}
\newtheorem{thm}[df]{Theorem}
\newtheorem{cor}[df]{Corollary}
\newtheorem{ex}[df]{Example}
\newtheorem{rem}[df]{Remark}
\def\N{\mathbbm N}
\def\Z{\mathbbm Z}
\def\C{\mathbbm C}
\def\id{\operatorname{id}}
%\def\card{\operatorname{card}}
\journal{arXiv.org}
\begin{document}
\begin{frontmatter}
\title{A nice acyclic matching on the nerve of the partition lattice}
\author[ruelle]{Ralf Donau}
\address[ruelle]{Fachbereich Mathematik, Universit\"at Bremen, Bibliothekstra\ss e 1, 28359 Bremen, Germany}
\ead{ruelle@math.uni-bremen.de}
\begin{abstract}
The author has already proven that the space $\Delta(\Pi_n)/G$ is homotopy equivalent to a wedge of spheres of dimension $n-3$ for all natural numbers $n\geq 3$ and all subgroups $G\subset S_1\times S_{n-1}$. We construct an $S_1\times S_{n-1}$-equivariant acyclic matching on $\Delta(\Pi_n)$ together with a description of its critical simplices. This is also a more elementary approach to determining the number of spheres. We also develop new methods for Equivariant Discrete Morse Theory by adapting the Patchwork Theorem and poset maps with small fibers from Discrete Morse Theory.
\end{abstract}
\begin{keyword}
Discrete Morse Theory\sep Regular trisp\sep Acyclic matching\sep Equivariant homotopy
\end{keyword}
\end{frontmatter}
\section{Introduction}
Let $n\geq 3$ and let $\Pi_n$ denote the poset consisting of all partitions of $[n]:=\{1,\dots,n\}$ ordered by refinement, such that the finer partition is the smaller partition. Let $\overline{\Pi}_n$ denote the poset obtained from $\Pi_n$ by removing both the smallest and greatest element, which are $\{\{1\},\dots,\{n\}\}$ and $\{[n]\}$, respectively. We consider $\overline{\Pi}_n$ as a category, which is acyclic, and define $\Delta(\overline{\Pi}_n)$ to the nerve of the acyclic category $\overline{\Pi}_n$, which is a regular trisp, see \cite[Chapter 10]{buch}. The symmetric group $S_n$ operates on $\Delta(\overline{\Pi}_n)$ in a natural way.

@@

In this paper we construct an $(S_1\times S_{n-1})$-equivariant acyclic matching on the face poset ${\cal F}(\Delta(\overline{\Pi}_n))$ of $\Delta(\overline{\Pi}_n)$ for $n\geq 3$ such that we have a description of the critical simplices. This induces an acyclic matching on $\Delta(\overline{\Pi}_n)/G$ for any subgroup $G\subset S_1\times S_{n-1}$.

@@

By Remark \ref{card} the cardinality of $C_n$ is $(n-1)!$ which equals the cardinality of $S_1\times S_{n-1}$. Hence the action is free.
\end{proof}
Let $G\subset S_1\times S_{n-1}$ be an arbitrary subgroup.
\begin{rem}
The cardinality of $C_n/G$ is the index of $G$ in $S_1\times S_{n-1}$.
\end{rem}
\begin{proof}
Apply Lemma \ref{freetrans}.
\end{proof}
Now Theorem \ref{rdthm} follows as a corollary. We can either apply Corollary \ref{moeglichkeit1} or Corollary \ref{moeglichkeit2}.
\begin{cor}
The topological space $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of spheres of dimension $n-3$. The number of spheres is the index of $G$ in $S_1\times S_{n-1}$.
\end{cor}
Now we consider the top cohomology $H^{n-3}(\Delta(\overline{\Pi}_n);\C)$ of $\Delta(\overline{\Pi}_n)$. For each simplex $\sigma\in C_n$ we obtain a $\sigma^*\in H^{n-3}(\Delta(\overline{\Pi}_n);\C)$ which is represented by a cocycle that maps $\sigma$ to $1$ and all other simplices to $0$. We obtain a basis $\{\sigma^*\mid\sigma\in C_n\}$ of the top cohomology which is called \emph{splitting basis} in \cite{wachs}.
\section*{Acknowledgments}
The author would like to thank Dmitry N. Kozlov for this interesting problem, Ragnar Freij and Giacomo d'Antonio for the helpful discussions. Furthermore the author would like to thank another anonymous reader of a previous version of this paper.
\bibliographystyle{model1-num-names}
\begin{thebibliography}{00}
\bibitem{arone}
G. Arone, \textit{A branching rule for partition complexes}, \texttt{arXiv:1508.05086 [math.AT]} (2015).
\bibitem{donau2}
R. Donau, \textit{On a quotient topology of the partition lattice with forbidden block sizes}, Topology and its Applications \textbf{159} (8) (2012), pp. 2052-2057.
\bibitem{donau}
R. Donau, \textit{Quotients of the order complex $\Delta(\overline{\Pi}_n)$ by subgroups of the Young subgroup $S_1\times S_{n-1}$}, Topology and its Applications \textbf{157} (16) (2010), pp. 2476-2479.
\bibitem{torsion}
R. Donau, \textit{Quotients of the topology of the partition lattice which are not homotopy equivalent to wedges of spheres}, \texttt{arXiv:1202.4368 [math.AT]} (2012).
\bibitem{forman}
R. Forman, \textit{Morse theory for cell complexes}, Adv. Math. \textbf{134} (1) (1998), pp. 90-145.
\bibitem{freij}
R. Freij, \textit{Equivariant discrete Morse theory}, Discrete Mathematics \textbf{309} (12) (2009), pp. 3821-3829.
\bibitem{hatcher}
A. Hatcher, \textit{Algebraic Topology}, Cambridge University Press, 2008.
\bibitem{clmap}
D.N. Kozlov, \textit{Closure maps on regular trisps}, Topology and its Applications \textbf{156} (15) (2009), pp. 2491-2495.
\bibitem{buch}
D. Kozlov, \textit{Combinatorial Algebraic Topology}, Algorithms and Computation in Mathematics \textbf{21}, Springer-Verlag Berlin Heidelberg, 2008.
\bibitem{wachs}
M.L. Wachs, \textit{On the (co)homology of the partition lattice and the free Lie algebra}, Discrete Mathematics \textbf{193} (1-3) (1998), pp. 287-319.
\end{thebibliography}
\end{document}
